---
title: "KPZ314 SDM Practical – Data-engineering workflow"
format: html
execute:
  echo: true
  message: false
  warning: false
---

```{r setup, include=FALSE}
library(targets)
tar_config_set(script = "_targets.R")

# auto‑build if the cache is empty
if (!tar_exist_objects(names = "pa_df")) {
  message("Building targets (~3‑5 min on fresh clone)…")
  tar_make()
}

pa_df        <- tar_read(pa_df)
grid_df      <- tar_read(grid_df)
fold_lookup  <- tar_read(folds)
pa_cv_df     <- tar_read(pa_cv_df)
```

> **Learning goals for this worksheet**

> LG‑1 Raw → PA dataset pipeline  
> LG‑2 Spatial cross-validation rationale  
> LG‑3 Visual QC of sampling bias

## 1  Snapshot of the raw ALA downloads

```{r}
library(dplyr); library(ggplot2)

pa_df %>%
  count(species, pa) %>% 
  tidyr::pivot_wider(names_from = pa,
                     values_from = n,
                     names_prefix = "pa_") %>% 
  knitr::kable()
```

> **TASK 1.1** Why do you think there are far fewer absences than presences?  
> Hint: we chose only 1 km grid cells that were occupied by at least 2 *other* species.

> **TASK 1.2** What class imbalance (pa_1 vs pa_0) would you flag as *severe?*  
> Give a numeric rule-of-thumb and justify it. Think about the species ecology.

## 2  Map presences & pseudo-absences

```{r}
pres_palette <- c("0" = "#D7191C",  "1" = "#2C7BB6")

ggplot(pa_df, aes(lon, lat, colour = factor(pa))) +
  geom_point(size = 0.8, alpha = 0.6) +
  coord_equal() +
  scale_colour_manual(values = pres_palette,
                      name = "PA flag",
                      labels = c("0" = "pseudo-absence",
                                 "1" = "presence")) +
  theme_minimal() +
  facet_wrap(~ species, ncol = 2) +
  labs(title = "ALA records snapped to 0.01° grid")
```

> **TASK 2.1** Visually, which species appears most clearly separated?  
> Explain what sampling biases might have led to this observed pattern.

## 3  Overlay an environmental predictor layer on Tasmania

```{r}
library(scales)

this_pred <- "human_footprint"

colnames(tar_read(grid_df)[, 3:17])
```

```{r}
ggplot(grid_df, aes(lon, lat, fill = .data[[this_pred]])) +
  geom_tile() +
  coord_equal() +
  theme_void() +
  scale_fill_viridis_c(option = "B", name = this_pred)
```

> **TASK 3.1** This was human footprint, with the highest intensity around urban areas. 
> Pick a different predictor from the list above (e.g., annprecip) and regenerate the map.  
> Which environmental gradient shows the clearest north–south trend?

## 4  Building spatial CV folds

```{r}
folds <- tar_read(folds)

# 1. Plot all points, coloured by fold
p <- ggplot(folds, aes(lon, lat, colour = factor(fold))) +
  geom_point(size = 1.2, alpha = 0.7) +
  scale_colour_viridis_d(name = "Fold", option = "turbo") +
  coord_equal() +
  theme_minimal()

# 2. Compute centroids of each fold cluster
centroids <- folds %>%
  group_by(fold) %>%
  summarise(lon = mean(lon), lat = mean(lat))

# 3. Overlay fold numbers in red at centroids
p + 
  geom_text(
    data = centroids,
    aes(lon, lat, label = fold),
    colour = "red", fontface = "bold", size = 3.5, show.legend = FALSE
  ) +
  labs(title = "Spatial CV folds (numbered by centroid)",
       x = "Longitude", y = "Latitude")
```

> **TASK 4.1** Why might random (non-spatial) CV over-estimate model accuracy  
> when evaluating SDMs?

> **TASK 4.2** We chose folds 5, 11, 19 and 23 for the held-out test set.  
> Why do you think these areas were selected? What is the trade-off at play here?

## 5  Train/test split check

```{r}
pa_cv_df %>%
  count(type, species, pa) %>%
  tidyr::pivot_wider(names_from = pa,
                     values_from = n,
                     names_prefix = "pa_") %>%
  knitr::kable()
```

> **TASK 5.1** Each species’ test set has ≥ ~100 presences. But some have few absences. 
> Explain why that might matters for how well we can train a good discriminating SDM.

## 6  Reflect

1. List two biases inherent in citizen-science data and the pipeline steps that
   mitigate them.  
2. Suggest an additional predictor that might improve model performance and
   justify its inclusion.
